---
title: "Vendi Scores"
output: github_document
vignette: >
  %\VignetteIndexEntry{vendi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Vendi Scores are a family of diversity metrics that are flexible, interpretable, and unsupervised.
Defined as the entropy associated with the eigenvalues of a sample similarity matrix $\mathbf{K}$, the Vendi Score 
only requires a pair-wise similarity measure. The Vendi Score is defined in [(Friedman and Dieng, 2023)](https://arxiv.org/abs/2210.02410). The order $q$ 
of the Vendi Score weight the importance of rare and common elements in the diversity computation, as described in this [paper](https://arxiv.org/abs/2310.12952). 

## Computing the Vendi Scores
```{r setup}
library(VendiScore)
```

To demonstrate some of the nice features of the Vendi Score, lets generate a few simple 1D mixture of Gaussian Datasets. 
```{r, fig.width=7, fig.height=4}


generate_data <- function(n_components=2, sd=0.1, tot_samp=360){
  set.seed(42)
  means <- rep(0, n_components)
  for(i in 1:length(means)){
    means[i] <- 3*(i-length(means))/2
  }
  data <- c()
  for (i in 1:n_components) {
    data <- c(data, rnorm(tot_samp / n_components, mean = means[i], sd = sd))
  }
  
  return(data)
}
par(mfrow = c(2, 3))
# First plot: 2 Gaussians
set.seed(42)
tot_samp <- 360
sd <- 0.1
data2 <- generate_data(2, sd=sd, tot_samp=tot_samp) 
hist(data2, main = "2 Gaussians", xlab = "Value", col = "lightblue", breaks = 10)

# Second plot: 3 Gaussians
data3 <- generate_data(3, sd=sd, tot_samp=tot_samp) 
hist(data3, main = "3 Gaussians", xlab = "Value", col = "lightblue", breaks = 10)

# Third plot: 4 Gaussians
data4 <- generate_data(4, sd=sd, tot_samp=tot_samp) 
hist(data4, main = "4 Gaussians", xlab = "Value", col = "lightblue", breaks = 10)

data5 <- generate_data(5, sd=sd, tot_samp=tot_samp) 
hist(data5, main = "5 Gaussians", xlab = "Value", col = "lightblue", breaks = 15)

data6 <- generate_data(6, sd=sd, tot_samp=tot_samp) 
hist(data6, main = "6 Gaussians", xlab = "Value", col = "lightblue", breaks = 15)
```

Now we can evaluate the behavior of the Vendi Score on these datasets. 

The Vendi Scores takes your data, a pair-wise similarity function, and the order $q$
as arguments. The data can be in any format that supports indexing. Here, we use a
simple Radial Basis Function (RBF) kernel to define similarity between pairs of samples.

```{r, fig.width=7}
rbf_kernel <- function(x, y, gamma = 0.75) exp(-gamma * sum((x - y)^2))
vendi_scores <- c(score(data2, rbf_kernel, q=1.), score(data3, rbf_kernel, q=1.), 
                  score(data4, rbf_kernel, q=1.), score(data5, rbf_kernel, q=1.), 
                  score(data6, rbf_kernel, q=1.))

x <- 2:6  # Shared x-axis

plot(x, vendi_scores, type = "l", col = "blue", lwd = 2,
     ylab = "Vendi Score", xlab = "Number of Gaussians", ylim = range(0,7), xaxt='n')
points(x, vendi_scores, col = "blue", pch = 16)
axis(1, at = seq(from = min(x), to = max(x), by = 1))
```

As we can see from above, the Vendi Score increases in direct correspondance with the increasing number of gaussians. 
